A Survey of Scientific Workflow Management Systems

### 2014-04-07 paper.tex
comments to sections

%Workflows considered:
% MOTEUR2
% Taverna
% Swift
% Pegasus
% Triana
% Galaxy
% Makeflow
% Kepler/Ptolemy
% Dryad
% Science pipes
% Spark
% Linda
% DaGue
% Wings

@@

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

@@

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field
The present work analyzes contemporary SWEs in order to find the capabilities
and key features in terms of the characteristics described in the introduction
part of this paper. The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work since we rely on the peer reviewed publications for information on a
given SWE. Some features, such as extendability of SWE strongly depend on the
extension points designed in the code; consequently, we consider it present in
only those SWEs that have a documented description or demonstration of the
given feature.

@@

%
\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the
University of North Carolina, Wilmington, facilitates a graphical composition
and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
Ptolemy II framework. GridNexus workflows are expressed in an XML-based
scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
customizable and extensible scripting language allowing users to write their
own workflow processor types. GridNexus workflows are grid oriented and can be
deployed on grid resources accessed through a persistent WS-RF (Web Services
Resource Framework) interface.

@@

% Put this somewhere else, either in the analysis or may be in the conclusions
% and summary.
%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language
%developed as a part of the current work is similar in spirit to the
%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A SwiftScript workflow
%contains statically typed datasets that can be specified using ``annotations''
%in the script. Unlike SwiftScript, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. SwiftScript
%provides for fault-tolerance and exception handling at run-time which is
%lacking currently in gscript. The underlying datasets organization in swift is
%represented using an XDTM XML-representation. This means that effectively, a
%workflow is SwiftScript plus the XDTM data representation. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. SwiftScript supports creation of nested-workflows in
%the form of compound procedures. Conditional and Loop execution is supported
%using the if and for/while statements in both the languages.

%
\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.

@@

>>>>

%Workflows considered:
% MOTEUR2
% Taverna
% Swift
% Pegasus
% Triana
% Galaxy
% Makeflow
% Kepler/Ptolemy
% Dryad
% Science pipes
% Spark
% Linda
% DaGue
% Wings
% PGRADE Portal

@@

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing
    %Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

@@

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field

% Say that different workflow systems have different design goals. It is hard
% to chose one from the existing or build your own. It is almost equally
% complex to build own just as it is complex to create own programming
% platform. We hope to provide by the way of this paper a guide for new adopters on how to chose a workflow language appropriate for their needs. We ask and answer the following questions of the workflow systems that we study in this paper:
%-- What were the original design goals?
%-- What are the main application areas that the system is currently being used?
%-- What are the characteristics and capabilities of the language? expressivity features?
%-- What are the characteristics and capabilities of the run-time? parallelism? concurrency? task and data grouping?
%-- What kinds of computing and computational infrastructure does the workflow system cater to? Grids, Clouds, Clusters, Supercomputers, fast IB networks? Accelerators?
%-- How have they evolved? What are the possible future directions?

The present work analyzes contemporary SWEs in order to find the capabilities
and key features in terms of the characteristics described in the introduction
part of this paper. The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work since we rely on the peer reviewed publications for information on a
given SWE. Some features, such as extendability of SWE strongly depend on the
extension points designed in the code; consequently, we consider it present in
only those SWEs that have a documented description or demonstration of the
given feature.

@@

%
\paragraph{Dryad/--}

@@

% Put this somewhere else, either in the analysis or may be in the conclusions
% and summary.
%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language
%is similar in spirit to the
%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A Swift script 
%contains statically typed datasets that can be declared
%in the script. Unlike Swift script, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
%represented using XML-representation. This means that effectively, a
%workflow is a DAG. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. Conditional and Loop execution is supported
%using the if and foreach statements in both languages.
%

\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.

@@

The GridNexus \cite{GridNexus} SWE, developed at the
University of North Carolina, Wilmington, facilitates a graphical composition
and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
Ptolemy II framework. GridNexus workflows are expressed in an XML-based
scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
customizable and extensible scripting language allowing users to write their
own workflow processor types. GridNexus workflows are grid oriented and can be
deployed on grid resources accessed through a persistent WS-RF (Web Services
Resource Framework) interface.
